Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2018-07-23 reviewers_response.md

Correct. Let's look into keeping the ERCC spike-ins in the object.

> 2b. It requires installation of the development branch of bioconductor, which impacts the usefulness of the workflow to the average user. I expect the authors will revise this tutorial when Bioconductor 3.6 is released and use of the devel branch is no longer necessary. Additionally `slingshot` is an requirement, but currently only exists on github and no SHA1 provided. I hope that `slingshot` will be added as a bioconductor package shortly. In the meantime, a tag must be added to the git repo for the release being used in this workflow and instructions provided for how to install this tag. Additionally, the authors may wish to note that installation instructions for the packages will be provided at the end of the workflow so that someone proceeding sequentially will not be tripped up.

Slingshot is in Bioconductor now. Good point about the installation instructions being at the end. I think we can provide a short section on installation at the beginning.

> 2c. Opaque code is presented in order to generate plots, e.g.
```{r}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20, xlab = "Component1", ylab = "Component2") legend(x = "topleft", legend = names(pal), cex = .5, fill = pal, title = "Sample")
```

> While this complexity may be necessary, perhaps some of it could be encapsulated as accessor functions in the package? Too much complexity here may cause users to miss the forest for the trees.

@@

The plotReduceDim() function now addresses this.

> The authors could better motivate (or at least explain the impact of) some of the default parameters and procedures.
> Why do we set a zcut threshold of 3 for the `scone` filtering? 
> Why K=50 for zinbwave?
> RSEC parameters

> How should a user decide on a value for these parameters?

@@

> "Note that, in this case, the low-dimensional matrix W is not included in the computation of residuals to avoid the removal of the biological signal of interest.": I understood this sentence only on a second pass through. One problem is that you haven't defined W in the text yet (only in Figure 4). I would only reference the matrix W if you have defined it.

We agree, that the reference to the model is too specific. We have modified the sentence. [TODO -- still not sure how to best describe this step]

> Figure 6: Can you change the figure width so that PC1 is not squished?

We actually dropped Figure 6 as using the low-dimensional signal from ZINB-WaVE (rather than PCA on the normalized counts) is the recommended way to visualize the results, and we thought that it was confusing for the reader to show PCA at this stage of the workflow.

@@

> Is there a circularity to the recovery of published clusters in Figure 6? Was ZINB-WaVE used in Fletcher (2017)?

ZINB-WaVE was not used in the original paper, which used a more complex normalization strategy, based on regressing out a set of QC measures. One key result in favor of the ZINB-WaVE projection is that with this method slingshot is able to estimate the correct lineages with minimal supervision (only the starting cluster), while in the original paper, one of the terminal states had to be specified.

DR: Make sure that with the original analysis we had to use a semi-supervised approach.

@@

> Can you say what the meaning of the color white is in Figure 8 (in the text or caption near this figure)?

We have added a better description of the Figure in the caption.

@@

> Figure 15 refers back to Figure 2 but does not use the same color scheme for the known cell types, so the reader cannot verify if you've recovered the lineages from the publication. It would be good therefore to have a legend for these figures (Fig 15 and following) indicating which cell types the colors refer to (this information is in the unlabeled table above, but should be included as a legend here).

Thanks for the suggestion, we agree that having a similar color scheme will help the reader make the connection between the results of Figure 15 and Figure 2. We have changed the colors to match as close as possible those of the published figure. Note that we do this right after the clustering step, so that all figures (from Figure 8 to Figure 17) have consistent colors.

@@

> Can you briefly describe what a GAM is ahead of Figure 17?

We have added a sentence describing GAMs ahead of Figure 17.

### 2018-07-23 workflow/workflow.Rmd
Addressing reviews

@@

```{r examinemakeConsensus, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package."}
plotClusters(dataObj)
```

DR: Is there a way to check the older results? Where do I find the saved objects?

@@

From the original published work, we know that the start cluster should correspond to HBCs and the end clusters to MV, mOSN, and mSUS cells. Additionally, we know that GBCs should be at a junction before the differentiation between MV and mOSN cells (Figure \@ref(fig:stemcelldiff)). The correspondance between the clusters we found here and the original clusters is as follows.

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | red | original 1, 5
c2 | mSUS | blue | original 4, 7
c3 | mOSN  | green | original 9, 12 
c4 | GBC | orange | original 2, 3, 11
c5 | Immature Neuron | purple | original 10, 14
c6 | MV | brown | original 15
c7 | mOSN | light blue | original 9

@@

More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time for each gene using the `gam` package. We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).

>>>>

The correspondance between the clusters we found here and the original clusters is as follows.

To facilitate the comparison between our results and the original clusters, we set a color scheme that allows a direct visual comparison with Figure \@ref(fig:stemcelldiff). We can change the clustering color scheme using `clusterExperiment`'s `recolorClusters` function.

```{r colors}
new_col <- col_clus[c("1", "4", "12", "3", "10", "15", "9")]
names(new_col) <- paste0("c", 1:7)
dataObj <- recolorClusters(dataObj, value = new_col)
```

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | green | original 1, 5
c2 | mSUS | magenta | original 4, 7
c3 | mOSN  | orange | original 9, 12 
c4 | GBC | cyan | original 2, 3, 11
c5 | Immature Neuron | purple | original 10, 14
c6 | MV | blue | original 15
c7 | mOSN | yellow | original 9

@@

```{r examinemakeConsensus, fig.cap="RSEC: Candidate clusterings found using the function RSEC from the clusterExperiment package. The top row corresponds to the consensus clustering and is color-coded accordingly. The other rows correspond to the individual clustering across which the consensus was computed. In each row, the color is chosen to match as close as possible the color of the cluster from the previous row with the largest overlap."}
plotClusters(dataObj, existingColors = "firstOnly")
```

@@

From the original published work, we know that the start cluster should correspond to HBCs and the end clusters to MV, mOSN, and mSUS cells. Additionally, we know that GBCs should be at a junction before the differentiation between MV and mOSN cells (Figure \@ref(fig:stemcelldiff)).

@@

More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time for each gene using the `gam` package.

Generalized Additive Models (GAMs) [@gam] are a class of models that replaces the usual linear function of the coefficients with a sum of smooth functions. In the code below, we use a special case of the GAM in which only one covariate (pseudotime) is present. This correspond to a simple loess regression, but with the added advantage of providing a framework for the testing of the significance of the association between pseudotime and expression.

We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
